Fail Faster

### 2025-10-08 paper/paper.tex
typo
meeting nits

%
We address the overhead of
many common generator abstractions using \textit{multi-stage
  programming} (or {\em staging}), a well-studied technique for
building fast DSLs "without regret"~\cite{lms}. Beyond the standard
staging transformations, we apply advanced staging techniques~\cite{partially-static,kovacs24, carrette05}
that let us leverage our domain-specific understanding of PBT generators to achieve the fastest possible code.
We refer to this approach as \name.
To demonstrate its generality, we implement staged generator DSLs in
two strict functional languages: OCaml and Scala 3.  In OCaml, we
build \camlname{}, based on \bq{}---a high-quality PBT library in
OCaml, authored by software engineers at the trading firm Jane Street~\cite{bq}.  In Scala 3, we build \scalaname{}, based on ScalaCheck---the
language's standard PBT library~\cite{sc}.

@@

In summary, we show that PBT generator DSLs incur significant
performance costs across languages and present two interventions that
significantly improve their efficiency while preserving their
idiomatic style.  Concretely, we offer the following contributions:
\begin{enumerate}
\item We identify two key sources of inefficiency in PBT generator
libraries---abstraction overhead and choice of \rand---both of which
significantly impact performance (Section~\ref{section:motiv}).
\item We present \name{}, a staging technique that eliminates the
abstraction overhead of generators. We apply \name{} to standard
generator libraries in both OCaml and Scala 3, showcasing its
generality (Section~\ref{section:faster-generators}).
\item We present a novel methodology for quantifying 
      the precise speedup achieved using staging and fast randomness,
      in which the performance benefit of each intervention can be 
      attributed to the interventions themselves, as opposed to 
      random chance (Section~\ref{section:eval}).
\item We demonstrate that writing generators using \name{} and fast
\rands{} yields substantial performance improvements, both separately
and in combination, and we show that these performance improvements
extend to significantly improved bug-finding speed (Section~\ref{section:eval}).
\end{enumerate}

@@

    let let_insert (cx : 'a code) : 'a code t =
      {cps = fun k -> k .< let x = .~cx in .~(k .<x>.) >.}
  end

>>>>

%
We address the overhead of
many common generator abstractions using \textit{multi-stage
  programming} (or {\em staging}), a well-studied technique for
building fast DSLs "without regret"~\cite{lms}. 
%
The simple monadic structure of typical PBT generators lend themselves naturally
to well-known staging transformations. Beyond these basic techniques, we observe
that common usage patterns of fundemental PBT-specific combinators---e.g., weighted choices with statically known options---are amenable
to advanced staging techniques~\cite{partially-static,kovacs24, carrette05}.

% Beyond the standard
% staging transformations, we apply advanced 
% that let us leverage our domain-specific understanding of PBT generators to achieve the fastest possible code.

We refer to this approach as \name.
To demonstrate its generality, we implement staged generator DSLs in
two strict functional languages: OCaml and Scala 3.  In OCaml, we
build \camlname{}, based on \bq{}---a high-quality PBT library in
OCaml, authored by software engineers at the trading firm Jane Street~\cite{bq}.  In Scala 3, we build \scalaname{}, based on ScalaCheck---the
language's standard PBT library~\cite{sc}.

@@

In summary, we show that PBT generator DSLs incur significant
performance costs across languages and present two interventions that
significantly improve their efficiency while preserving their
idiomatic style.  Concretely, we offer the following contributions:
\begin{enumerate}
\item We identify two key sources of inefficiency in PBT generator
libraries---abstraction overhead and choice of \rand---both of which
significantly impact performance (Section~\ref{section:motiv}).
\item We present \name{}, a staging technique that eliminates the
abstraction overhead of generators. We apply \name{} to standard
generator libraries in both OCaml and Scala 3, showcasing its
generality (Section~\ref{section:faster-generators}).
% \item We present a novel methodology---enabled by the pointwise semantic equivalence of original and \name{}-optimized generators---for quantifying 
%       the precise speedup achieved using staging and fast randomness,
%       in which the performance benefit of each intervention can be 
%       attributed to the interventions themselves, as opposed to 
%       random chance (Section~\ref{section:eval}).
\item We demonstrate that writing generators using \name{} and fast
\rands{} yields substantial performance improvements, both separately
and in combination, and we show that these performance improvements
extend to significantly improved bug-finding speed (Section~\ref{section:eval}).
\end{enumerate}

@@

    let let_insert (cx : 'a code) : 'a code t =
      {cps = fun k -> .< let x = .~cx in .~(k .<x>.) >.}
  end
